Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support).

### 2017-02-08 csf-2017/paper.tex
Make a start on presenting the capability machine.

% Domains
\newcommand{\plaindom}[1]{\mathrm{#1}}
\newcommand{\Caps}{\plaindom{Cap}}
\newcommand{\Words}{\plaindom{Word}}
\newcommand{\Addrs}{\plaindom{Addr}}
\newcommand{\ExecConfs}{\plaindom{ExecConf}}
\newcommand{\RegName}{\plaindom{RegisterName}}
\newcommand{\Regs}{\plaindom{Reg}}
\newcommand{\Heaps}{\plaindom{Mem}}
\newcommand{\Mems}{\Heaps}
%\newcommand{\HeapSegments}{\plaindom{MemSegment}}
\newcommand{\HeapSegments}{\plaindom{MemSeg}}
\newcommand{\MemSegments}{\HeapSegments}
\newcommand{\Confs}{\plaindom{Conf}}
\newcommand{\Instrs}{\plaindom{Instructions}}
\newcommand{\nats}{\mathbb{N}}
\newcommand{\ints}{\mathbb{Z}}
\newcommand{\Perms}{\plaindom{Perm}}
\newcommand{\Globals}{\plaindom{Global}}

@@

@@

\section{Introduction}
\label{sec:introduction}
\todo[inline]{See the latex comments for the next paragraphs!}

@@

In this paper, we make progress towards the goal of efficient, provably
watertight enforcement of encapsulation and control-flow integrity on capability
machines.  Specifically, we make the following contributions:
\begin{itemize}
\item We formalise a simple but representative capability machine featuring
  local capabilities much like CHERI's, and define its operational semantics.
\item We define a novel stack management discipline for enforcing control-flow
  integrity and encapsulation of local state on the stack. The discipline relies
  solely on local capabilities (used to represent stack and return pointers) and
  does not require OS support in the form of a trusted shadow stack or special
  call and return instructions. It also does not require separate per-component
  stacks and instead works with a single, fairly standard contiguous stack. It
  supports higher-order cross-component calls (e.g. (potentially)
  cross-component function pointers) and can be efficient assuming one
  additional piece of processor support: an instruction for efficiently clearing
  a range of memory.
\item We provide a theorem (capability-safety) that is (by far) the most general
  and powerful formulation of the guarantees offered by a capability machine.
  The theorem also expresses the specific guarantees offered by the machine for
  local capabilities. It is very general and not tied to a specific ABI or a
  specific way of using the system's capabilities.
\item The above theorem is formulated in terms of a unary step-indexed logical
  relation with recursively defined Kripke worlds. It is inspired by one used by
  \cite{Devriese:2016ObjCap} for capability-safety in a JavaScript-like lambda
  calculus, but has two novel technical characteristics. The natural CPS style
  of assembly code with return pointers as continuations is modeled using a
  \emph{single} orthogonal closure, rather than the biorthogonal closure that
  has been previously used for lambda calculi. Perhaps more surprisingly, we
  show that the special nature of local capabilities can be elegantly expressed
  using a variant of an idea that has been used by
  \cite{Dreyer:2010:IHS:1863543.1863566} to express well-bracketedness
  guarantees in a lambda calculus: Kripke worlds with a weak and a strong future
  world relations. Interestingly, the two future world relations generally work
  very similarly to Dreyer et al.'s, except that they turn up in quite a
  different place in the logical relation.
\item We demonstrate the power of our results by applying them to challenging
  examples, specifically constructed to demonstrate local state encapsulation
  and control-flow integrity guarantees in the presence of cross-component
  function pointers. The examples demonstrate both the power of our formulation
  of capability-safety and our stack management discipline.
\end{itemize}

@@

\subsection{Introduce capability machines}
\subsection{Enforcing properties on a capability machine}

Privacy of local state using memory capabilities.
Encapsulation of private state using enter pointers or CHERI's ccall.
(mention in passing secure compilation, full abstraction)

\subsection{Well-bracketed control flow and encapsulation of stack frames}

Hard to enforce using standard capabilities:
\begin{itemize}
\item Well-bracketed control flow can be broken if the adversary stores return pointers.
(alternative: trusted stack)
\item Encapsulation of local stack frames can be broken if the adversary stores stack pointers.
\end{itemize}

\subsection{Local capabilities}

\begin{itemize}
\item Introduced by CHERI for this purpose.
\item Explain how they can be used for passing return pointers and stack pointers.
\item But this is not watertight.
\item Example attack:
\begin{itemize}
\item In a first invocation, adversary fills the entire unused part of the stack with copies of the current stack pointer
\item In a second invocation, adversary can load the old stack pointer of the stack
\item He now has access to private state that was pushed on the stack in between.
\end{itemize}
\item So:
\begin{itemize}
\item Can we make it (provably) watertight?
\item Is the trusted stack necessary? Why not just pass return pointers locally?
\item General question: How to reason about capability machines and
local capabilities?
\end{itemize}
\end{itemize}

\subsection{Contributions}

\begin{itemize}
\item Formal model of a simple but representative capability machine with local capabilities
\item Detailed study of how to do safe stack and return pointer management in this setting, taking into account:
\begin{itemize}
\item untrusted adversary
\item higher-order code: callbacks passed to and received from the adversary
\item efficient stack management
\item no OS support
\end{itemize}
\item Logical relation for reasoning about code in this capability machine
\item Fundamental theorem that expresses the guarantees offered by a capability machine for untrusted code
\item Technical contributions: reuse existing ideas in a new way:
\begin{itemize}
\item replace biorthogonal closure by a single orthogonal closure
because assembly languages remove the distinction between the
continuation and the arguments
\end{itemize}
\end{itemize}
LB: I think we should have some discussion of the flexibility /
  strength of our LR.  The LR defines which computations we think of
  as well-behaved.   
 We should give simple / trivial examples of
  code not in the LR. But we may also want to emphasize that the LR is 
  not too restricting, e.g., it does not enforce a certain calling
  scheme (examplied by Example \verb!f1! which
  does not use the stack and the following examples which do use the
  stack). This ties in to the discussion of biorthogonality, see the
  following paragraphs.

LB: we need to relate this carefully to Hur-Dreyer, who used
      biorthogonality even though they also worked with an assembly
      language. (If I recall correctly, they assumed some properties
      of the low level language, which is why the biortho was the
      right ting, but we need to check.)

LS: It is my impression that they were interested in the relation
between high-level and low-level programs. They are therefore in
particular interested in low-level programs compiled from high-level
programs which means that the continuation is always invoked in the 
same way. Our realisation was that anything we pass as an argument 
in a "call" can be used as a continuation, so the continuation 
relation was redundant. Even though we do try to give our programs
some structure using the macros, our logical relation is strong enough
to handle unstructured programs as well.
\begin{itemize}
\item STSs with public/private transitions for dealing with local capabilities: play the same role as before, but in different places of the LR.
\end{itemize}
\begin{itemize}
\item Demonstrate all of this on several examples:
\begin{itemize}
\item security examples
\item the most challenging examples from existing literature on reasoning about well-bracketed control flow in lambda calculi
\item some compartmentalisation result?
\item whatever else we do..
\end{itemize}
\end{itemize}

\subsection{Outline}

\section{A capability machine with local capabilities}

Formalisation

@@

\section{Stack and return pointer management using local capabilities}

\begin{itemize}
\item Issues to consider:
\begin{itemize}
\item no global store-local capabilities
\item clearing the stack
\item clearing (some) registers
\item constructing return pointers
\item constructing callbacks
\item invoking the adversary's callbacks
\end{itemize}
\item Reusable macro-instructions for stack and return pointer management
\begin{itemize}
\item taking care of the stuff above
\end{itemize}
\end{itemize}

@@

\section{Logical relation}
In order to be able to reason about programs running on the capability
machine and prove properties such as well-bracketedness and
encapsulation about them, we define a unary logical relation.

% Intuition
The logical relation is a semantic model of all the words on the
capability machine that are definitely safe to give to an
adversary. Whether a capability is safe to give to an adversary
depends heavily on the contents of the memory since instructions as well
as capabilities may reside in memory. Hence we index the logical
relation over worlds, which are used to model the memory, and our logical
relation is thus an example of what is often referred to as a Kripke
logical relation. A world consists of a finite set of regions, and each region
can be understood as a protocol describing the possible evolution of
part of the memory. Memory changes during an
execution since new memory can be allocated and part of the existing
memory can change. We use two future world relations to capture
such changes to the memory in our model.
%

>>>>

% Domains
\newcommand{\plaindom}[1]{\mathrm{#1}}
\newcommand{\Caps}{\plaindom{Cap}}
\newcommand{\Words}{\plaindom{Word}}
\newcommand{\Addrs}{\plaindom{Addr}}
\newcommand{\ExecConfs}{\plaindom{ExecConf}}
\newcommand{\RegName}{\plaindom{RegName}}
\newcommand{\Regs}{\plaindom{Reg}}
\newcommand{\Heaps}{\plaindom{Mem}}
\newcommand{\Mems}{\Heaps}
%\newcommand{\HeapSegments}{\plaindom{MemSegment}}
\newcommand{\HeapSegments}{\plaindom{MemSeg}}
\newcommand{\MemSegments}{\HeapSegments}
\newcommand{\Confs}{\plaindom{Conf}}
\newcommand{\Instrs}{\plaindom{Instructions}}
\newcommand{\nats}{\mathbb{N}}
\newcommand{\ints}{\mathbb{Z}}
\newcommand{\Perms}{\plaindom{Perm}}
\newcommand{\Globals}{\plaindom{Global}}

@@

\newcommand{\isdef}{\mathrel{\overset{\makebox[0pt]{\mbox{\normalfont\tiny\sffamily def}}}{=}}}
\newcommand\bnfdef{\mathrel{::=}}

@@

\section{Introduction}
\label{sec:introduction}

@@

In this paper, we make progress towards the goal of efficient, provably
watertight enforcement of encapsulation and control-flow integrity on capability
machines.  Specifically, we make the following contributions:
\begin{itemize}
\item We formalise a simple but representative capability machine featuring
  local capabilities much like CHERI's, and define its operational semantics.
\item We define a novel stack management discipline for enforcing control-flow
  integrity and encapsulation of local state on the stack. The discipline relies
  solely on local capabilities (used to represent stack and return pointers) and
  does not require OS support in the form of a trusted shadow stack or special
  call and return instructions. It also does not require separate per-component
  stacks and instead works with a single, fairly standard contiguous stack. It
  supports higher-order cross-component calls (e.g. (potentially)
  cross-component function pointers) and can be efficient assuming one
  additional piece of processor support: an instruction for efficiently clearing
  a range of memory.
\item We provide a theorem (capability-safety) that is (by far) the most general
  and powerful formulation of the guarantees offered by a capability machine.
  The theorem also expresses the specific guarantees offered by the machine for
  local capabilities. It is very general and not tied to a specific ABI or a
  specific way of using the system's capabilities.
\item The above theorem is formulated in terms of a unary step-indexed logical
  relation with recursively defined Kripke worlds. It is inspired by one used by
  \cite{Devriese:2016ObjCap} for capability-safety in a JavaScript-like lambda
  calculus, but has two novel technical characteristics. The natural CPS style
  of assembly code with return pointers as continuations is modeled using a
  \emph{single} orthogonal closure, rather than the biorthogonal closure that
  has been previously used for lambda calculi. Perhaps more surprisingly, we
  show that the special nature of local capabilities can be elegantly expressed
  using a variant of an idea that has been used by
  \cite{Dreyer:jfp12} to express well-bracketedness
  guarantees in a lambda calculus: Kripke worlds with a weak and a strong future
  world relations. Interestingly, the two future world relations generally work
  very similarly to Dreyer et al.'s, except that they turn up in quite a
  different place in the logical relation.
\item We demonstrate the power of our results by applying them to challenging
  examples, specifically constructed to demonstrate local state encapsulation
  and control-flow integrity guarantees in the presence of cross-component
  function pointers. The examples demonstrate both the power of our formulation
  of capability-safety and our stack management discipline.
\end{itemize}

@@

% \subsection{Contributions}

% \begin{itemize}
% \item Formal model of a simple but representative capability machine with local capabilities
% \item Detailed study of how to do safe stack and return pointer management in this setting, taking into account:
% \begin{itemize}
% \item untrusted adversary
% \item higher-order code: callbacks passed to and received from the adversary
% \item efficient stack management
% \item no OS support
% \end{itemize}
% \item Logical relation for reasoning about code in this capability machine
% \item Fundamental theorem that expresses the guarantees offered by a capability machine for untrusted code
% \item Technical contributions: reuse existing ideas in a new way:
% \begin{itemize}
% \item replace biorthogonal closure by a single orthogonal closure
% because assembly languages remove the distinction between the
% continuation and the arguments
% \end{itemize}
% \end{itemize}
% LB: I think we should have some discussion of the flexibility /
%   strength of our LR.  The LR defines which computations we think of
%   as well-behaved.   
%  We should give simple / trivial examples of
%   code not in the LR. But we may also want to emphasize that the LR is 
%   not too restricting, e.g., it does not enforce a certain calling
%   scheme (examplied by Example \verb!f1! which
%   does not use the stack and the following examples which do use the
%   stack). This ties in to the discussion of biorthogonality, see the
%   following paragraphs.

% LB: we need to relate this carefully to Hur-Dreyer, who used
%       biorthogonality even though they also worked with an assembly
%       language. (If I recall correctly, they assumed some properties
%       of the low level language, which is why the biortho was the
%       right ting, but we need to check.)

% LS: It is my impression that they were interested in the relation
% between high-level and low-level programs. They are therefore in
% particular interested in low-level programs compiled from high-level
% programs which means that the continuation is always invoked in the 
% same way. Our realisation was that anything we pass as an argument 
% in a "call" can be used as a continuation, so the continuation 
% relation was redundant. Even though we do try to give our programs
% some structure using the macros, our logical relation is strong enough
% to handle unstructured programs as well.

% DD: I think the above comment by LS is correct: the reason why the biortho is
% the right thing for Hur-Dreyer is that their LR only accepts well-behaved
% programs which treat the return pointer as a continuation (i.e. don't store
% it), while ours is more general.

% \begin{itemize}
% \item STSs with public/private transitions for dealing with local capabilities: play the same role as before, but in different places of the LR.
% \end{itemize}
% \begin{itemize}
% \item Demonstrate all of this on several examples:
% \begin{itemize}
% \item security examples
% \item the most challenging examples from existing literature on reasoning about well-bracketed control flow in lambda calculi
% \item some compartmentalisation result?
% \item whatever else we do..
% \end{itemize}
% \end{itemize}

\subsection{Outline}
\todo[inline]{TODO}

\section{A capability machine with local capabilities}
\begin{figure}
  \begin{align*}
    \addr   &\in& \Addrs &\isdef \nats\\
    w &\in&\Words &\isdef \ints + \Caps \\
    \perm   &\in& \Perms &::= \noperm \mid \readonly\mid \readwrite\mid \readwritel\mid \exec\mid \entry\mid \rwx\mid \rwlx\\
    \gl&\in&\Globals & ::= \glob \mid \local \\
     && \Caps  &::= \left\{
                       \begin{multlined}
                         ((\perm,\gl),\start,\addrend,\addr)\mid\\
                         \start\in\Addrs,\addrend \in
                         \Addrs\cup \{\infty\}
                       \end{multlined} \right\}\\
    r       &\in& \RegName&\bnfdef \pcreg\mid r_0\mid r_1\mid\ldots\\
    \reg &\in& \Regs  &\isdef \RegName \rightarrow \Words\\
    m&\in& \Heaps &\isdef \Addrs \rightarrow \Words \\
    \Phi    &\in& \ExecConfs  &\isdef \Regs \times \Heaps \\
    &&\Confs &::= \ExecConfs + \{\failed \} + \{\halted\} \times \Heaps \\
    \ms     &\in& \MemSegments &::= \Addrs \parfun \Words
  \end{align*}

  \begin{array}{rcl}
    \lv    &::=& \refreg{r} \\
    \hv    &::=& \refheap{r}\\
    \rv    &::=& n \mid \lv \\
    i      &::=& 
                 \jmp{\lv} \mid 
                 \jnz{\lv}{\rv} \mid
                 \move{\lv}{\rv} \mid 
                 \load{\lv}{\hv} \mid 
                 \store{\hv}{\rv} \mid  \\
           &   & \plus{\lv}{\rv}{\rv} \mid 
                 \lea{\lv}{\rv} \mid 
                 \restricttwo{\lv}{\rv} \mid 
                 \subseg{\lv}{\rv}{\rv} \mid  \\
           &   & \isptr{\lv}{\rv} \mid 
                 \getp{\lv}{\lv} \mid 
                 \getl{\lv}{\lv} \mid 
                 \getb{\lv}{\lv} \mid
                 \gete{\lv}{\lv} \mid\\
           &   & \geta{\lv}{\lv} \mid 
                 \fail \mid
                 \halt 
  \end{array}
  \caption{The syntax of our capability machine assembly language.}
  \label{fig:syntax}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}[main node/.style={}]
    \node[main node] (7) {$\rwlx$};
    \node[main node] (8) [below left of=7] {$\readwritel$};
    \node[main node] (1) [below right of=7] {$\rwx$};
    \node[main node] (2) [below right of=1] {$\exec$};
    \node[main node] (3) [below right of=2] {$\entry$};
    \node[main node] (4) [below left of=1] {$\readwrite$};
    \node[main node] (5) [below right of=4] {$\readonly$};
    \node[main node] (6) [below right of=5] {$\noperm$};

    \path[every node/.style={font=\sffamily\small}]
    (7) edge (8)
    (7) edge (1)
    (8) edge (4)
    (1) edge (2)
    (2) edge (3)
    (2) edge (5)
    (3) edge (6)
    (1) edge (4)
    (4) edge (5)
    (5) edge (6);
  \end{tikzpicture}

  \caption{Permission hierarchy}
  \label{fig:perm-hier}
\end{figure}

We define the syntax of our capability machine in Figure~\ref{fig:syntax}. We
assume an infinite set of addresses $\Addrs$ and define machine words as either
integers or capabilities of the form $((\perm,\gl),\start,\addrend,\addr)$. Such
a capability represents the authority to execute permissions $\perm$ on the
memory array $[\start,\addrend]$, together with a current address $\addr$ and a
tag $\gl$ which indicates whether the capability is global or local. The
available permissions and their ordering are depicted in
Figure~\ref{fig:perm-hier}: the permissions include null permission ($\noperm$),
readonly ($\readonly$), read/write ($\readwrite$), read/execute ($\exec$) and
read/write/execute ($\rwx$) permissions. Additionally, there are somewhat
special read/write-local ($\readwritel$), read/write-local/execute ($\rwlx$) and
enter ($\entry$) permissions which we will explain below.

We assume a set of register names $\RegName$. We define register files $\reg$ as
functions mapping register names to words and memories as functions that map
addresses to words. The state of the entire machine is represented as a
configuration that is either a running $\Phi \in \ExecConfs$ containing a memory
and a register file, or a failed or halted state, where the latter keeps track
of the final state of memory.

@@

\section{Stack and return pointer management using local capabilities}
\begin{itemize}
\item Issues to consider:
\begin{itemize}
\item no global store-local capabilities
\item clearing the stack
\item clearing (some) registers
\item constructing return pointers
\item constructing callbacks
\item invoking the adversary's callbacks
\end{itemize}
\item Reusable macro-instructions for stack and return pointer management
\begin{itemize}
\item taking care of the stuff above
\end{itemize}
\end{itemize}

@@

\section{Logical relation}
In order to be able to reason about programs running on the capability
machine and prove properties such as well-bracketedness and
encapsulation about them, we define a unary logical relation.
% Intuition
The logical relation is a semantic model of all the words on the
capability machine that are definitely safe to give to an
adversary. Whether a capability is safe to give to an adversary
depends heavily on the contents of the memory since instructions as well
as capabilities may reside in memory. Hence we index the logical
relation over worlds, which are used to model the memory, and our logical
relation is thus an example of what is often referred to as a Kripke
logical relation. A world consists of a finite set of regions, and each region
can be understood as a protocol describing the possible evolution of
part of the memory. Memory changes during an
execution since new memory can be allocated and part of the existing
memory can change. We use two future world relations to capture
such changes to the memory in our model.
%
